Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-26 paper.tex
Updates from Overleaf

For XB, all samples trace back to a originating node which is the product of a recombination between a B.1 sample (specifically the strain England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of the genome from the start up to a maximum position of 23604bp), and two other UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of 27389bp) and B.1.36.8 (the remaining fragment of the genome). As described below, in the long ARG the equivalent recombination node has only 2 parents, with no involvement of  B.1.36.8, and it is possible that the third parent in the wide ARG is artifactual. An added complication is that in both ARGs, additional non-X-designated samples such as descend from the not only a small number of are six further recombination nodes nested within the XB ``clade''. However, five are ancestral to only negligible fractions of the designated XB samples. The sixth accounts for about 17\% of the designated XB nodes, and involves a recombination between descendants of the originating XB recombination. More specifically strain USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination between an XB grandparent and its XB grandchild. This could well be artifactual.

@@

\includegraphics[width=0.4\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
\\ % &
% \multicolumn{2}{c}{
\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}
% }
\end{tabularx}
\caption{\label{fig:pango-single-origin-graph}
Single origin Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown back to the nearest sampled viral genome, with dotted lines showing continuation of the ARG. Nodes are coloured by Nextclade Pango designation; smaller nodes are not samples and their Pango designation was imputed where possible. (B) Details for the single-origin Pango X lineages in the long ARG. (C) XAG origins in the long ARG.}
\end{figure}

@@

[Discuss panel C where XAG shares a common recombinant origin with samples from several other X lineages]
All XAG samples share a single origin; however this is also shared with all XAA samples, and includes some, but not all, of the samples identified as XAB, XQ, and XU by Nextclade. Note that if the GISAID designations are used, many of the samples marked here as XAB are reclassified as BA.2, and XAG becomes fully monophyletic (supplementary figure \ref{fig:pango_XAG_gisaid_graph}; this also gives identities of the samples in the graph)

@@

In Figure XXX we give the details of these multiple pango lineages, along with an example lineage subgraph.

@@

As well as cases where a single recombination event explains several X lineages (such as XAG above), many of the Pango X lineages involve more complex origins, in which samples which have not been previously identified as recombinants also descend from originating recombination events. In addition these complex topologies often include extra recombination nodes within the topology that descends from the main originating event. An example is XB in which recombination between a B.1 sample and B.1.627 sample leads both to the XB samples and to a "hairball" of further recombination nodes whose descendants have not been identified as recombinants by Nextclade (blue nodes, Figure~\ref{fig:complex_origins_graph})  ...

@@

@@

>>>>

For XB, all samples trace back to a originating node which is the product of a recombination between a B.1 sample (specifically the strain England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of the genome from the start up to a maximum position of 23604bp), and two other UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of 27389bp) and B.1.36.8 (the remaining fragment of the genome). As described later, in the long ARG the equivalent recombination node has only 2 parents, with no involvement of  B.1.36.8, and it is possible that the third parent in the wide ARG is artifactual. Additional complications arise because additional non-X-designated samples such as B.1.634 also descend from this recombination (also true in the long ARG, see below), and there are also a small number of further recombination nodes nested within the XB ``clade''. However, all but one of these nested recombinations are ancestral to only a negligible fractions of the designated XB samples. The exception accounts for about 17\% of the designated XB nodes, and involves a recombination between descendants of the originating XB recombination. More specifically strain USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination between an XB grandparent and its XB grandchild. This could well be artifactual.

@@

\includegraphics[width=0.4\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
\\ % &
% \multicolumn{2}{c}{
\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}
% }
\end{tabularx}
\caption{\label{fig:pango-single-origin-graph}
Single origin Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown outwards to the nearest sampled viral genome; dotted lines show where the ARG continues. Nodes are coloured by Nextclade Pango designation; smaller symbols are nodes inserted by  sc2ts (i.e. not samples); their Pango designation has been imputed where possible. (B) Details for the single-origin Pango X lineages in the long ARG. (C) XAG origins in the long ARG.}
\end{figure}

@@

[Discuss panel C where XAG shares a common recombinant origin with samples from several other X lineages]
All XAG samples share a single origin; however this is also shared with all XAA samples, and includes some, but not all, of the samples identified as XAB, XQ, and XU by Nextclade. Note that if the GISAID designations are used, many of the samples marked here as XAB are reclassified as BA.2, and XAG becomes fully monophyletic (supplementary figure \ref{fig:pango_XAG_gisaid_graph}; this also gives identifying labels to the sample nodes, which can be converted to EPI\_ISL indentifiers using Supplemeentary table XXX )

@@

In Figure XXX we give the details of these multiple pango lineages, along with an example lineage subgraph. As with the subgraphs in Figure~\ref{tab:pango-single-origin}, the equivalent XD plot with GISAID Pango designations, along with identifying labels, is shown for comparison in Supplementary Figure XXX

@@

As well as cases where a single recombination event explains several X lineages (such as XAG above), many of the Pango X lineages involve more complex origins, in which samples which have not been previously identified as recombinants also descend from originating recombination events. In addition these complex topologies often include extra recombination nodes within the topology that descends from the main originating event. An example is XB in the long ARG, plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination between a B.1 sample and B.1.627 sample leads both to the XB samples and to a "hairball" of further recombination nodes whose descendants have not been identified as recombinants by Nextclade (blue nodes, "pre-Alpha" VoC). However, in this subgraph, there is only a single sample, identified as XB by Nextclade, which is descended from these
nested recombination events; moreover, this sample is not identified as XB by GISAID (Supplementary Figure~\ref{}). We suggest that investigating cases of complex origins, and identifing which (if any) of the nested, further recombination events may be artifactual, is an important area of future research.

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XA_gisaid_large_graph.pdf}
\caption{\label{fig:pango_XA_gisaid_graph}  DESCRIBE ME}
\end{figure}

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XD_gisaid_large_graph.pdf}
\caption{\label{fig:pango_XD_gisaid_graph}  DESCRIBE ME}
\end{figure}

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XB_gisaid_large_graph.pdf}
\caption{\label{fig:pango_XB_gisaid_graph}  DESCRIBE ME}
\end{figure}
